---
title: "p8105_hw2_zlb2008"
author: "Zakari L. Billo"
date: "2025-09-29"
output: html_document
---

```{r setup, echo=FALSE}
library(tidyverse)
library(readxl)
```

## Problem 1

First, clean the data in pols-month.csv. Use separate() to break up the variable `mon` into integer variables `year`, `month`, and `day`; replace month number with month name; create a `president` variable taking values `gop` and `dem`, and remove `prez_dem` and `prez_gop`; and remove the day variable.

```{r}
pols_month_df <-
  read_csv('data/pols-month.csv', na = c("NA", ".", "")) |>
  janitor::clean_names() |>
  separate(mon, 
    into = c("year", "month", "day"),
    sep = "-"
  ) |>
  mutate(month = case_match(
    month, 
    "01" ~ "January",
    "02" ~ "February",
    "03" ~ "March",
    "04" ~ "April",
    "05" ~ "May",
    "06" ~ "June",
    "07" ~ "July",
    "08" ~ "August",
    "09" ~ "September",
    "10" ~ "October",
    "11" ~ "November",
    "12" ~ "December"
    ),
    president = case_match(
    prez_gop, 
    1 ~ "gop",
    .default = case_match(prez_dem, 1 ~ "dem")
    )) |>
  select(-prez_dem, -prez_gop, -day)
```

Second, clean the data in snp.csv using a similar process to the above. For consistency across datasets, arrange according to year and month, and organize so that year and month are the leading columns.

```{r}
snp_df <-
  read_csv('data/snp.csv', na = c("NA", ".", "")) |>
  janitor::clean_names() |>
  separate(date, 
    into = c("month", "day", "year"),
    sep = "/"
  ) |>
  mutate(month = case_match(
    month, 
    "1" ~ "January",
    "2" ~ "February",
    "3" ~ "March",
    "4" ~ "April",
    "5" ~ "May",
    "6" ~ "June",
    "7" ~ "July",
    "8" ~ "August",
    "9" ~ "September",
    "10" ~ "October",
    "11" ~ "November",
    "12" ~ "December"
    )) |>
  select(year, month, everything(), -day)
```

Third, tidy the unemployment data so that it can be merged with the previous datasets. This process will involve switching from “wide” to “long” format; ensuring that key variables have the same name; and ensuring that key variables take the same values.

```{r}
unemployment_df <-
  read_csv('data/unemployment.csv', na = c("NA", ".", "")) |>
  janitor::clean_names() |>
  pivot_longer(
    jan:dec,
    names_to = "month", 
    values_to = "perc_unemployment"
  ) |>
  select(year, month, perc_unemployment)
```

Join the datasets by merging snp into pols, and merging unemployment into the result.

```{r}
snp_pols_df = 
  left_join(pols_month_df, snp_df, by = c("year", "month"))

```


## Including Plots


